A weakly structured stem for human origins in Africa

### 2023-01-18 paper.tex
basal population split

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S, then we could observe disproportionate
ancestry contributions from different stems in contemporary groups. We observed
two merger events. The first, between Stem 1S and Stem 2, resulted in the
formation of an ancestral Khoe-San population $\sim120$ka. The second event
$\sim100$ka between Stem 1E and Stem 2, resulted in the formation of the
ancestors of eastern/western Africans including the ancestors of
``Out-of-Africans''.
% The following sentence may need clarification
While $120$ka can be interpreted as the basal human population divergence,
we suggest conceptualizing this event as the time of most recent
shared ancestry among sampled populations.
``Population divergence'' in population genetics is a complex concept due
to the co-estimation of divergence time and subsequent migration;
methods assuming clean vs reticulated splits can infer different split dates
(Figure~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
Therefore, wide variation exists in estimates of divergence time in the
literature\citep{Bergstrom2021-iw,Henn2018-rf}.

>>>>

A population reticulation involves multiple stems contributing genetically to
the formation of a group. One way in which this can happen is through the
geographic expansion of one or both stems. For example, if during MIS 5, either
Stem 1S (Figure~\ref{fig:best-fit-models}B) from southern Africa moved
northward thereby encountering Stem 2, or Stem 2 moved from central/western
Africa southward into Stem 1S, then we could observe disproportionate
ancestry contributions from different stems in contemporary groups. We observed
two merger events. The first, between Stem 1S and Stem 2, resulted in the
formation of an ancestral Khoe-San population $\sim120$ka. The second event
$\sim100$ka between Stem 1E and Stem 2, resulted in the formation of the
ancestors of eastern/western Africans including the ancestors of
``Out-of-Africans''.
% The following sentence may need clarification
Reticulated models do not have a unique and 
well-defined "basal" human population divergence.
We suggest conceptualizing the events $120$ka as the time of most recent
shared ancestry among sampled populations.
However, interpreting population divergence times in population genetics is 
always challenging due
to the co-estimation of divergence time and subsequent migration;
methods assuming clean vs reticulated splits can infer different split dates
(Figure~\ref{fig:supp-rccr-data} and Figure~\ref{fig:supp-IM-misspecification}).
Therefore, wide variation exists in estimates of divergence time in the
literature\citep{Bergstrom2021-iw,Henn2018-rf}.
